---
title: "Causal Inference and Regression"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(knitr)
set.seed(04132021)
```

Causal inference can be characterized as a predictive problem, where the question is _what would have happened under different circumstances._

\vfill

Simulate and visualize data with two potential outcomes

\vfill

```{r}
n <- 40
y0 <- rnorm(n)
y1 <- y0 + rnorm(n, mean = 2)
```

```{r}
dat <- tibble(id = factor(rep(1:n, 2)), response = c(y0, y1), 
              `potential outcome` = rep(c('y0','y1'), each = n))

dat %>% ggplot(aes(y = response, x = `potential outcome`)) + 
  geom_violin(draw_quantiles = c(.025, .5, .975)) + 
  geom_jitter(aes(color = id),  shape = rep(c(0:19),4)) +
  theme_bw() + theme(legend.position = 'none') +
  ggtitle('Potential Outcomes')
```

\newpage

Then randomly assign treatments to each unit and visualize differences
\vfill

```{r}
treatment <- sample(rep(0:1, each = n/2))
sample_dat <- tibble(id = factor(1:n), response = y0, 
                     `potential outcome` = rep('y0', n)) %>% filter(treatment == 0) %>% bind_rows(
                    tibble(id = factor(1:n), response = y1, 
                     `potential outcome` = rep('y1', n)) %>% filter(treatment == 1)   
                     )
```

\vfill

```{r}
sample_dat %>% ggplot(aes(y = response, x = `potential outcome`)) + 
  geom_violin(draw_quantiles = c(.025, .5, .975)) + 
  geom_jitter(aes(color = id)) + theme_bw() + theme(legend.position = 'none') +
    ggtitle('Factual Outcomes')

```


\newpage

#### Pre-treatment covariates

Consider a setting with: 

- pre-treatment covariates for sampling units (either continuous or categorical)

\vfill

- treatments (control + treatment) applied to the sampling units.

\vfill

Begin with a randomized block design (paired comparisons, is a special case with blocks of size 2). Analysis can 

\vfill
- use differences between treatment and control

\vfill
- use differences between treatment and control for each (block)

\vfill
- Adjusting for pre-treatment variables (categorical/blocks or continuous)

$$y_i =  \tau z_i + X_i \beta + \epsilon_i,$$
where $z_i$ is an indicator for treatment and $\tau$ is the average treatment in the grade.

\vfill

Note this assumes the average treatment is the same for each block. How would we allow for varying treatment effects?
