---
title: "Causal Inference and Regression"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(knitr)
set.seed(04152021)
```


Recall, we can adjust for pre-treatment variables (categorical/blocks or continuous)

$$y_i =  \tau z_i + X_i \beta + \epsilon_i,$$
where $z_i$ is an indicator for treatment and $\tau$ is the average treatment effect.

\vfill

Even if a blocking design isn't used, pre-treatment variables can be used to account for differences between the control and treatment groups.

\vfill

or in a hierarchical framework (with random effects) as

$$y_i =  \tau z_i + \theta_{j[i]} + \epsilon_i,$$
where $\theta_j \sim N(0, \sigma^2_\theta)$ is a random effect centered at zero.

```{r, eval = F}
stan_glmer(response ~ treatment + (1|block), data = dat)
```

\vfill

Generally, blocks are included as random effects if they are not directly of interest, such as individuals. However, if the blocking factor is of interest, such as locations with sun vs. shade in a plot, fixed effects are common.

\vfill

While these model specifications control for blocks, both assume constant treatment effects across groups.

\vfill

Interactions (or varying treatment effects through hierarchical models) can be used to capture different treatment levels.

\newpage

Stratification, which is similar to blocking, is a technique to divide a population into groups and estimate totals within each strata. Then population totals can be computed by combining the strata with appropriate sampling weights.

\vfill

_Post-stratification_ uses the same idea of estimating totals within groups, but doesn't require (need?) to have groups pre-defined as part of the sampling approach. Post-stratification can be used to both estimate totals with groups, and, adjust for differences between the proportion of each class in the sample and population (using weights).

\vfill

The post stratification estimates are also conditional average treatment effects.